---
title: "Chapter 3 - Discrete Probability Distribution"
author: "Dr. Alessandro Ruggieri"
affiliation: University of Nottingham
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
    fig_width: 4
    fig_height: 3
    fig_caption: true
  html_document:
    toc: yes
    toc_depth: '3'
---


### Binomal distribution
A random variable distributed as a Binomial(n,p) has a density equal to $p(x) = C(n, x) p^x (1-p)^{(n-x)}$ for $x>0$. The mean is equal to E(x)=$np$ while the variance is Var(x)=$np (1 − p)$

We can compute the density p(x) using the function dbinom( )

```{r}
# trials
n<-10
# probability
p<-0.3
# number of success
x<-3
# density at x (probability of x successes out of n trial)
dbinom(x,n, p)
```

We can also compute the distribution function P(x) using the function pbinom( )

```{r}
# distribution function at x  (probability of up to x successes out of n trial)
pbinom(x, n, p)
```

Alternatively, we can also compute the distribution function by summing the densities accordingly, i.e.

```{r}
# distribution function at x = 3
out<-0
i<-0
while(i<= 3) {
  dist<-dbinom(i,n, p)
  out = dist +out
  i<-i+1
}
out
```

We can use the function pbinom( ) to compute the probability that number of successes lie between two values, a and b.

```{r}
# lower bound - minimum number of successes
a<-2
# upper bound - maximum number of successes
b<-5
# probability number of successes is between and and bout of n trial
sum(dbinom(a:b, n, p))
```

or alternative, we can use the function pbinom( ) and subtract the distribution evaluated at a-1 to the distribution evaluated at b, i.e
```{r}
# probability number of successes is between a and b out of n trial
pbinom(b, n, p) - pbinom(a-1, n, p) 
```


Notice that the Bernulli distribution can be understood as a Binomial for the case of n=1 trial.

If we want to draw a k numbers from a Binomial distribution B(n,p), we can use the function rbinom( ):
```{r}
# numbers to draw
k<-10000
# draw k numbers
y<-rbinom(k, n, p)  
```

and the we can display the density using a histogram:
```{r}
# histogram
hist(y, main="Probability density of Binomial r.v.")
```


### Poisson distribution
A random variable distributed as a Poisson($\lambda$) has density equal to $p(x) = \lambda^x \frac{\exp(-\lambda)}{x!}$ for $x\geq 0$. 

The mean and variance are E(x) = Var(x) = $\lambda$. We can compute the density p(x) using the function dpois( )

```{r}
# rate of success
lambda<-3
# number of success
x<-3
# density at x (probability of x successes in interval of time with rate of success lambda)
dpois(x,lambda)
```

We can also compute the distribution function P(x) using the function ppois( )

```{r}
# distribution function at x  (probability of up to x successess in interval of time with rate of success lambda)
ppois(x,lambda)
```

If we want to draw a k numbers from a Binomial distribution B(n,p), we can use the function rbinom( )
```{r}
# numbers to draw
k<-10000
# draw k numbers
z<-rpois(k, lambda)  
```

and the we can display the density using a histogram:
```{r}
# histogram
hist(z, main="Probability density of Poisson r.v.")
```